Dynamical modeling of tidal streams

### 2014-01-02 tex/ms.tex
streams: first full draft pre-spell-check

%BOVY:NOTES:
% - GO through citations and be a little more reluctant to cite Sanders
% - do more justice to Eyre & Binney work
% - discuss hotter streams somewhere
% - say more about why dynammical large-scale modeling doesn't care about the small-scale structure
% - what about spherical potentials? dO/dJ has zero eigenvalue
% - We/I distinction
\documentclass[12pt,preprint]{aastex}
%\documentclass{emulateapj}
\usepackage{amssymb,amsmath}
%\usepackage{caption}
%\DeclareCaptionLabelSeparator{dotemdash}{.--- }
%\captionsetup[figure]{labelformat=simple, labelsep=dotemdash}
%\usepackage{color,hyperref}
% hypertex insanity
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\newcounter{address}
\setcounter{address}{1}
%\usepackage{sidecap}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\Eqnname}{Equation}
\newcommand{\equationname}{\eqnname}
\renewcommand{\tablename}{Table}
\renewcommand{\figurename}{Figure}
\newcommand{\figurenames}{\figurename~s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\normal}{\ensuremath{\mathcal{N}}}
\newcommand{\flag}[1]{\texttt{\lowercase{#1}}}
\newcommand{\feh}{\ensuremath{[\mathrm{Fe/H}]}}
\newcommand{\afe}{\ensuremath{[\alpha\mathrm{/Fe}]}}
\newcommand{\logg}{log g}
\newcommand{\Ro}{\ensuremath{R_0}}

@@

However, streams are observed in position--velocity coordinates and
for action--angle descriptions to be useful, we must be able to
calculate the transformation between these coordinate systems
efficiently. Until now this has required accurate phase--space data
\citep{Sanders13b} and specialized algorithms to calculate
action--angle coordinates that break down for the eccentric orbits on
which tidal-stream progenitors are typically found (that is, radial
and/or vertical actions of similar size as the angular momentum;
\citealt{Sanders12a}). However, even with Gaia, future stream data
will typically have large uncertainties compared to the intrinsic
dispersion of the stream and in particular the line-of-sight
velocities of the faintest stream members will not be observed for
many stars. Additionally, streams are superimposed on a non-negligible
background of field stars and background contamination cannot be
easily taken into account in any of the current stream-fitting
methods.

@@

Thus, in action--angle coordinates, the dynamics of tidal-stream
formation is simplified to linear growth of small initial angle
differences due to small action/frequency differences. The angle
direction $\unitvec{\veco} \equiv \Delta \veco / | \Delta \veco|$
along which a star moves away from the progenitor is related to the
initial action offset by $\Delta \veco \approx (\partial \veco /
\partial \vecj) \Delta \vecj = (\partial^2 H / \partial \vecj \partial
\vecj) \Delta \vecj$, in the limit of small action offsets, because
$\veco = \partial H / \partial \vecj$, where $H$ is the
Hamiltonian. For an approximately one-dimensional stream to form, the
Hessian matrix $(\partial^2 H / \partial \vecj \partial
\vecj)\large|_{\vecj^p}$ has to be dominated by a single large
eigenvalue; in this case the angle difference of stars stripped from
the stream at any time will fall along approximately the same
direction $\unitvec\veco$, and the stream will be essentially one
dimensional. In detail, because the distribution of initial offsets
$\Delta \vecj$ is not isotropic, the direction $\unitvec \veco$ is not
just determined by the potential (through $H$) and the progenitor
orbit, but also by the distribution of $\Delta \vecj$, which is set
through a combination of the internal dynamics of the progenitor and
the stripping process (see discussion in \citealt{Sanders13b}). For
dynamical modeling we can be agnostic about the relative importance of
all of these contributors to $\unitvec \veco$ by flexible modeling of
the distribution of $\Delta \vecj$ or $\Delta \veco$.

@@

A generative model of a tidal stream in frequency--angle space
requires three ingredients: (a) a model for the distribution of times
at which stars are stripped from the progenitor, (b) a prescription
for the distribution of frequency offsets from the progenitor at every
given stripping time, and (c) a description of the angle offsets for
any given frequency offset and stripping time. Having specified these
ingredients, we can then generate tidal debris at any given stripping
time and evolve it forward using the simple linear dynamics in
frequency--angle space discussed in
\sectionname~\ref{sec:dynamics}. The initial angle offsets are small
enough that they are likely unobservable even with futuristic data
(especially in the direction of the frequency offset, where the
initial angle offset is quickly overwhelmed by the subsequent
dynamical evolution; for the simulated stream used here this happens
after approximately $20\Myr$). For that reason, we will assume that
the initial angle offsets are independent of both stripping time and
frequency offset. In what follows, we will model them using a simple
isotropic Gaussian distribution.

@@

@@

The discussion of the simulated stream's properties as a function of
stripping time in the previous section indicates that we can build
simple models of the structure of tidal streams when considering them
in frequency--angle space. In this section, we propose a very simple
fiducial model that we will use in the remainder of this paper. This
model combines simple analytic forms for ease of computation that are
nevertheless realistic enough to provide adequate models for tidal
streams in many applications.

First, we approximate the bursty, non-uniform distribution of
stripping times in \figurename~\ref{fig:gd1_times} with a uniform
distribution
\begin{align}\label{eq:pt}
  \begin{split}
    p(\ts) & \propto 1\,,\qquad 0 < \ts < t_d\\
    & = 0\,,\qquad \mathrm{otherwise}\,.
  \end{split}
\end{align}
The thick, solid histogram in the right panel of
\figurename~\ref{fig:gd1_times} shows a coarser-binned histogram of
stripping times, indicating that the uniform distribution is a good
approximation over intervals larger than a radial period ($\approx
400\Myr$). The \emph{disruption time} $t_d$ is a free parameter of the
model that needs to be determined for each stream. For the model of
the simulated stream used here we set $t_d = 4.5\Gyr$.

Similarly, we model the distribution of frequency offsets between
stream particles and the progenitor to be independent of stripping
time, even though the middle panel of \figurename~\ref{fig:gd1_times}
clearly shows that particles are stripped at larger frequency
differences at pericentric passages than between them. The fact that
there is no long-term trend in the typical frequency offset means that
this is a good model on time-scales larger than the radial
period. While this means that the small-scale structure of the stream
will not be perfectly represented by the model, the large-scale
structure of the stream will be fine. For using tidal streams to
constrain the gravitational potential, large arcs are most important,
such that this simple model will be adequate for such inferences.

The distribution of frequencies is three dimensional. Because we model
the leading and the trailing arm separately, a single-peaked
distribution suffices to describe the frequency distribution. A
general description in terms of a Gaussian distribution would have
nine free parameters. However, we can model it with fewer free
parameters as follows. First, following \citet{Eyre11a}, we
approximate the \emph{action} distribution as a Gaussian with standard
deviations
\begin{equation}
  \sigma_{J_R} = \frac{1}{\pi}\,\sigv\,\left(r_{\mathrm{apo}}-r_{\mathrm{peri}}\right)\,,\quad 
  \sigma_{L_Z} = \sigv\,r_{\mathrm{peri}}\,,\quad 
  \sigma_{J_Z} = \frac{2}{\pi}\,\sigv\,Z_{\mathrm{max}}\,,
\end{equation}
where $\sigv$ is the velocity dispersion of the progenitor,
$r_{\mathrm{apo}}$ and $r_{\mathrm{peri}}$ are the apo- and pericenter
radii of the progenitor orbit, respectively, and $Z_{\mathrm{max}}$ is
the maximum $Z$ reached on this orbit. We assume that the all
correlations between the actions are zero (see
\figurename~\ref{fig:gd1_jt}). Then we propagate this Gaussian
variance to frequency space using the Hessian $(\partial \veco /
\partial \vecj)\large|_{\vecj^p}$\footnote{The Hessian $(\partial
  \veco / \partial \vecj)\large|_{\vecj^p}$ can be computed using the
  method in \appendixname~\ref{sec:aa} by computing $\partial
  (\vecj,\veca) / \partial (\vecx,\vecv)$ and $\partial (\veco,\veca)
  / \partial (\vecx,\vecv)$ and forming $\partial \veco / \partial
  \vecj = \left(\partial (\vecj,\veca) / \partial
  (\vecx,\vecv)\right)^{-1}\,\left(\partial (\veco,\veca) / \partial
  (\vecx,\vecv)\right)$.}, which gives the variance matrix
$V_\veco$. The principal eigenvector of this matrix is the model
direction along which the stream spreads. While the model I propose
here is a simple one, it works well without having been tweaked for
the simulation under scrutiny. The angle between the principal
eigenvector of $V_\veco$ and the progenitor's frequency vector is
$0.50^\circ$ while the value measured from the simulation is
$0.18^\circ$. For a model with an isotropic action distribution, the
misalignment would be $1.28^\circ$.

@@

We then further constrain the mean frequency offset to lie along the
principal eigenvector of $V_\veco$. This mean frequency offset can be
described in terms of a parameter $\mu_\Omega = \Delta \Omega^m /
\sigma_{\Omega,1}$, where $\sigma_{\Omega,1}$ is the square root of
the largest eigenvalue of $V_\veco$ and $\Delta \Omega^m$ is the
one-dimensional mean frequency offset. The sign of this parameter sets
whether we are modeling a leading or a trailing
arm. \figurename~\ref{fig:gd1_dohist} shows the distribution of
frequency offsets in the direction of the median frequency offset in
the stream. The dashed line is a Gaussian fit to this distribution,
while the solid line is a fit of the form
\begin{equation}\label{eq:do}
  p(|\Delta \veco|_\parallel) = |\Delta \veco|_\parallel\,\mathcal{N}\left(|\Delta \veco|_\parallel|\Delta \Omega^m,\sigma_{\Omega,1}^2\right)\,,
\end{equation}
where $\mathcal{N}(\cdot|\cdot,\cdot)$ represents a Gaussian
distribution. The fit is equally good (because the dispersion is much
smaller than the mean) and we choose the second form because it
simplifies some of the stream-track calculations below. The best-fit
$(\Delta \Omega^m,\sigma_{\Omega,1})$ is $(0.19,0.033)\Gyr\inv$ in
both cases. Based on this information, a model with $\sigv =
0.365\kms$ and $\mu_\Omega = 6$ provides a good fit to the simulation
data. The distribution of frequency offsets in the two-dimensional
space perpendicular to the principal eigenvector of $V_\veco$ is
modeled as a zero-mean Gaussian with variance matrix given by the
projection of $V_\veco$ onto this space.

Because we expect the typical $\Delta\veco$ to scale with the
velocity dispersion of the progenitor (see also \citealt{Sanders13a}),
I conjecture that a constant $\mu_\Omega$ can be used for modeling
cold tidal streams of any (small-ish) velocity dispersion. Further
modeling of simulated streams, however, is necessary to check this and
to find the optimizal value of $\mu_\Omega$. Fixing $\mu_\Omega$
leaves a single free parameter \sigv\ for modeling the frequency
offset distribution of a tidal stream. In modeling to observed data,
this parameter needs to be fit and we expect it to be proportional to
the velocity dispersion of the progenitor, although this
proportionality needs to be checked more carefully (but see Figure 3
in \citealt{Sanders13a}, which shows that the size of the frequency
distribution scales as mass$^{1/3}$, or approximately as the velocity
dispersion). Similarly, the initial angle distribution, modeled here
as an isotropic Gaussian, has a charactaristic spread $\sigma_{\theta}$
which we set here to $0.003$ ($=\sigv/[122\kms]$), and we likewise
expect this spread to scale with the velocity dispersion of the
progenitor.

The fiducial model for a tidal stream in frequency--angle coordinates
proposed here is therefore described by essentially two free
parameters in addition to the progenitor's phase-space position: the
progenitor's velocity dispersion \sigv\ and the disruption time
$t_d$. As shown below, these two parameters are the most important in
determining the track of the stream. More sophisticated models can
leave $\mu_\Omega$ and $\sigma_{\theta}$ as free parameters as
well. In what follows, we will fix these to the values given in the
previous paragraph. It must be stressed, however, that all of these
parameters here have been fixed ``by-eye'', and better fits might be
possible using more quantitative fit procedures.

@@

First, we compute the distribution of $\Delta \opar$, the frequency
offset between stream members and the progenitor given the angle
offset $\Delta \apar$ as follows (considering $\Delta \opar$ to be
positive)
\begin{equation}\label{eq:poparapargeneral}
\begin{split}
  p(\Delta \opar|\Delta \apar) & = \int \dd \ts\,p(\Delta \opar,\ts|\Delta \apar)\, \\
  & \propto \int \dd \ts \,p(\Delta \apar | \Delta \opar,\ts)\,p(\Delta \opar,\ts)\,,\\
  & \approx \int \dd \ts \,\delta(\Delta \apar - \Delta \opar\,\ts)\,p(\Delta \opar,\ts)\,,\\
  & = \frac{1}{\Delta \opar}\,p(\Delta \opar,\frac{\Delta \apar}{\Delta \opar})\,,
\end{split}
\end{equation}
where in the penultimate step we have approximated the initial angle
distribution as a delta function (as the initial angle offset is small
with respect to the final offset). This is a fully general
expression. For the fiducial stream model, this can be simplified to
\begin{equation}\label{eq:poparapar}
\begin{split}
  p(\Delta \opar|\Delta \apar) \propto &\ \frac{1}{\Delta \opar}\,p(\Delta \opar)\,,\qquad 0 < \frac{\Delta \apar}{\Delta \opar} < t_d\,.\\
  & 0\,,\qquad \mathrm{otherwise}\,.
\end{split}
\end{equation}
\Eqnname~(\ref{eq:poparapar}) shows why we chose to model $p(\Delta
\opar)$ as a Gaussian multiplied with $|\Delta \opar|$. $p(\Delta
\opar|\Delta \apar)$ is then a Gaussian with mean $\Delta \Omega^m$
and variance $\sigma^2_{\Omega,1}$, for $\Delta \opar > \Delta \apar /
t_d$, and zero otherwise. The mean and variance of such a Gaussian is
straightforward to calculate in terms of error functions. The top left
panel of \figurename~\ref{fig:gd1_apar} shows the distribution of
$|\Delta \opar|$ versus $|\Delta \apar|$ (absolute value in order to
show the trailing and leading stream together) for the simulated
stream, as well as the mean and dispersion calculated from the
fiducial model. The dashed line shows $\Delta \apar = \Delta \Omega^m
\,t_d$, approximately the angle where the finite disruption time
starts to influence the mean $\Delta \opar$ of the stream. The finite
disruption time, $\approx 4.5\Gyr$ in this case, of a tidal stream
means that for stream members to have reached very large angle
differences with respect to the progenitor, they must have been
removed at large frequency differences. Around $\Delta \Omega^m
\,t_d$, average stream stars, \ie, those removed with the average
frequency offset, do not have a large enough offset to reach large
angle offsets. Therefore, even though the average frequency offset
does not change over time (see \figurename~\ref{fig:gd1_times}), the
average frequency offset, and therefore the average orbit, does change
with distance from the progenitor. It is in this sense that streams do
not follow a single orbit. This is discussed further in
\sectionname~\ref{sec:discussorbit}.

@@

We can calculate $p(\ts | \Delta \apar)$ as follows
\begin{equation}\label{eq:pdtapar}
\begin{split}
  p(\ts | \Delta \apar) & = \int \dd \Delta \opar\,p(\ts,\Delta \opar|\Delta \apar)\,,\\
  & = \int \dd \Delta \opar\,p(\ts|\Delta \opar,\Delta \apar)\,p(\Delta \opar|\Delta \apar)\,,\\
  & = p\left(\Delta \opar = \frac{\Delta \apar}{\ts}\right)\,\left|\frac{\Delta \apar}{(\ts)^2}\right|\,,
\end{split}
\end{equation}
where in the last step we have again approximated the initial
distribution of parallel angle offsets as a delta function. The first
factor in this equation is given by the expression given in
\equationname~(\ref{eq:poparapargeneral}) (in general) or in
\equationname~(\ref{eq:poparapar}) for the fiducial model. The lower
right panel of \figurename~\ref{fig:gd1_apar} shows the distribution
of stripping times for members of the simulated stream as well as the
mean and dispersion calculated from
\equationname~(\ref{eq:pdtapar}). Similar to the distribution of
$\Delta \opar$ as a function of $\Delta \apar$, the distribution of
$\ts$ has a kink at the angle offset that can only be reached by stars
that have to have been stripped at large frequency offsets (that is,
larger than average).

>>>>

%BOVY:NOTES:
% - We/I distinction
\documentclass[12pt,preprint]{aastex}
%\documentclass{emulateapj}
\usepackage{amssymb,amsmath}
%\usepackage{caption}
%\DeclareCaptionLabelSeparator{dotemdash}{.--- }
%\captionsetup[figure]{labelformat=simple, labelsep=dotemdash}
%\usepackage{color,hyperref}
% hypertex insanity
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\newcounter{address}
\setcounter{address}{1}
%\usepackage{sidecap}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\Eqnname}{Equation}
\newcommand{\equationname}{\eqnname}
\renewcommand{\tablename}{Table}
\renewcommand{\figurename}{Figure}
\newcommand{\figurenames}{\figurename~s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\normal}{\ensuremath{\mathcal{N}}}
\newcommand{\flag}[1]{\texttt{\lowercase{#1}}}
\newcommand{\feh}{\ensuremath{[\mathrm{Fe/H}]}}
\newcommand{\afe}{\ensuremath{[\alpha\mathrm{/Fe}]}}
\newcommand{\logg}{log g}
\newcommand{\Ro}{\ensuremath{R_0}}

@@

However, streams are observed in position--velocity coordinates and
for action--angle descriptions to be useful, we must be able to
calculate the transformation between these coordinate systems
efficiently. Until now this has required accurate phase--space data
and specialized algorithms to calculate action--angle coordinates that
break down for the eccentric orbits on which tidal-stream progenitors
are typically found (that is, radial and/or vertical actions of
similar size as the angular momentum; \citealt{Sanders12a}). However,
even with Gaia, future stream data will typically have large
uncertainties compared to the intrinsic dispersion of the stream and
in particular the line-of-sight velocities of the faintest stream
members will not be observed for many stars. Additionally, streams are
superimposed on a non-negligible background of field stars and
background contamination cannot be easily taken into account in any of
the current stream-fitting methods.

@@

Thus, in action--angle coordinates, the dynamics of tidal-stream
formation is simplified to linear growth of small initial angle
differences due to small action/frequency differences. The angle
direction $\unitvec{\veco} \equiv \Delta \veco / | \Delta \veco|$
along which a star moves away from the progenitor is related to the
initial action offset by $\Delta \veco \approx (\partial \veco /
\partial \vecj) \Delta \vecj = (\partial^2 H / \partial \vecj \partial
\vecj) \Delta \vecj$, in the limit of small action offsets, because
$\veco = \partial H / \partial \vecj$, where $H$ is the
Hamiltonian. For an approximately one-dimensional stream to form, the
Hessian matrix $(\partial^2 H / \partial \vecj \partial
\vecj)\large|_{\vecj^p}$ has to be dominated by a single large
eigenvalue; in this case the angle difference of stars stripped from
the stream at any time will fall along approximately the same
direction $\unitvec\veco$, and the stream will be essentially one
dimensional. In detail, because the distribution of initial offsets
$\Delta \vecj$ is not isotropic, the direction $\unitvec \veco$ is not
just determined by the potential (through $H$) and the progenitor
orbit, but also by the distribution of $\Delta \vecj$, which is set
through a combination of the internal dynamics of the progenitor and
the stripping process. For dynamical modeling we can be agnostic about
the relative importance of all of these contributors to $\unitvec
\veco$ by flexible modeling of the distribution of $\Delta \vecj$ or
$\Delta \veco$.

@@

A generative model of a tidal stream in frequency--angle space
requires three ingredients: (a) a model for the distribution of times
at which stars are stripped from the progenitor, (b) a prescription
for the distribution of frequency offsets from the progenitor at every
given stripping time, and (c) a description of the angle offsets for
any given frequency offset and stripping time. Having specified these
ingredients, we can then generate tidal debris at any given stripping
time and evolve it forward using the simple linear dynamics in
frequency--angle space discussed in
\sectionname~\ref{sec:dynamics}. The initial angle offsets are small
enough that they are likely unobservable even with futuristic data
(especially in the direction of the frequency offset, where the
initial angle offset is quickly overwhelmed by the subsequent
dynamical evolution; for the simulated stream used here this happens
after approximately $20\Myr$). For that reason, I will assume that the
initial angle offsets are independent of both stripping time and
frequency offset. In what follows, we will model them using a simple
isotropic Gaussian distribution.

@@

We can model streams in frequency space because the mapping $\vecj
\rightarrow \veco$ is of full rank for axisymmetric and triaxial
potentials. However, for spherical potentials this mapping only has
rank two, because $\Omega_\phi = \mathrm{sgn}(L_Z)\,\Omega_Z$, such
that the determinant $|\partial \veco / \partial \vecj|$ is zero. For
spherical potentials, tidal streams only spread in a five-dimensional
subspace and in modeling stream in spherical potentials we need to
keep the direction perpendicular to this subspace constant.

@@

The discussion of the simulated stream's properties as a function of
stripping time in the previous section indicates that we can build
simple models of the structure of tidal streams when considering them
in frequency--angle space. In this section, I propose a very simple
fiducial model that I will use in the remainder of this paper. This
model combines simple analytic forms for ease of computation that are
nevertheless realistic enough to provide adequate models for tidal
streams in many applications.

First, we approximate the bursty, non-uniform distribution of
stripping times in \figurename~\ref{fig:gd1_times} with a uniform
distribution
\begin{align}\label{eq:pt}
  \begin{split}
    p(\ts) & \propto 1\,,\qquad 0 < \ts < t_d\\
    & = 0\,,\qquad \mathrm{otherwise}\,.
  \end{split}
\end{align}
The thick, solid histogram in the right panel of
\figurename~\ref{fig:gd1_times} shows a coarser-binned histogram of
stripping times, indicating that the uniform distribution is a good
approximation over intervals larger than a radial period ($\approx
400\Myr$). The \emph{disruption time} $t_d$ is a free parameter of the
model that needs to be determined for each stream. For the model of
the simulated stream used here I set $t_d = 4.5\Gyr$.

Similarly, we model the distribution of frequency offsets between
stream particles and the progenitor to be independent of stripping
time, even though the middle panel of \figurename~\ref{fig:gd1_times}
clearly shows that particles are stripped at larger frequency
differences at pericentric passages than between them. The fact that
there is only a small long-term trend in the typical frequency offset
means that this is a good model on time-scales larger than the radial
period. While this means that the small-scale structure of the stream
will not be perfectly represented by the model, the large-scale
structure of the stream will be fine. For using tidal streams to
constrain the gravitational potential, large arcs are most important,
such that this simple model will be adequate for such inferences.

The distribution of frequencies is three dimensional. Because we model
the leading and the trailing arm separately, a single-peaked
distribution suffices to describe the frequency distribution. A
general description in terms of a Gaussian distribution would have
nine free parameters. However, we can model it with fewer free
parameters as follows. First, following \citet{Eyre11a}, we
approximate the \emph{action} distribution as a Gaussian with standard
deviations
\begin{equation}
  \sigma_{J_R} = \frac{1}{\pi}\,\sigv\,\left(r_{\mathrm{apo}}-r_{\mathrm{peri}}\right)\,,\quad 
  \sigma_{L_Z} = \sigv\,r_{\mathrm{peri}}\,,\quad 
  \sigma_{J_Z} = \frac{2}{\pi}\,\sigv\,Z_{\mathrm{max}}\,,
\end{equation}
where $\sigv$ is the velocity dispersion of the progenitor,
$r_{\mathrm{apo}}$ and $r_{\mathrm{peri}}$ are the apo- and pericenter
radii of the progenitor orbit, respectively, and $Z_{\mathrm{max}}$ is
the maximum $Z$ reached on this orbit. We assume that all of the
correlations between the actions are zero (see
\figurename~\ref{fig:gd1_jt}). Then we propagate this Gaussian
variance to frequency space using the Hessian $(\partial \veco /
\partial \vecj)\large|_{\vecj^p}$\footnote{The Hessian $(\partial
  \veco / \partial \vecj)\large|_{\vecj^p}$ can be computed using the
  method in \appendixname~\ref{sec:aa} by computing $\partial
  (\vecj,\veca) / \partial (\vecx,\vecv)$ and $\partial (\veco,\veca)
  / \partial (\vecx,\vecv)$ and forming $\partial \veco / \partial
  \vecj = \left(\partial (\vecj,\veca) / \partial
  (\vecx,\vecv)\right)^{-1}\,\left(\partial (\veco,\veca) / \partial
  (\vecx,\vecv)\right)$.}, which gives the variance matrix
$V_\veco$. The principal eigenvector of this matrix is the model
direction along which the stream spreads. While the model I propose
here is a simple one, it works well without having been tweaked for
the simulation under scrutiny. The angle between the principal
eigenvector of $V_\veco$ and the progenitor's frequency vector is
$0.50^\circ$ while the value measured from the simulation is
$0.18^\circ$. For a model with an isotropic action distribution, the
misalignment would be $1.28^\circ$.

@@

I then further constrain the mean frequency offset to lie along the
principal eigenvector of $V_\veco$. This mean frequency offset can be
described in terms of a parameter $\mu_\Omega = \Delta \Omega^m /
\sigma_{\Omega,1}$, where $\sigma_{\Omega,1}$ is the square root of
the largest eigenvalue of $V_\veco$ and $\Delta \Omega^m$ is the
one-dimensional mean frequency offset. The sign of this parameter sets
whether we are modeling a leading or a trailing
arm. \figurename~\ref{fig:gd1_dohist} shows the distribution of
frequency offsets in the direction of the median frequency offset in
the stream. The dashed line is a Gaussian fit to this distribution,
while the solid line is a fit of the form
\begin{equation}\label{eq:do}
  p(|\Delta \veco|_\parallel) = |\Delta \veco|_\parallel\,\mathcal{N}\left(|\Delta \veco|_\parallel|\Delta \Omega^m,\sigma_{\Omega,1}^2\right)\,,
\end{equation}
where $\mathcal{N}(\cdot|\cdot,\cdot)$ represents a Gaussian
distribution. The fit is equally good (because the dispersion is much
smaller than the mean) and I choose the second form because it
simplifies some of the stream-track calculations below. The best-fit
$(\Delta \Omega^m,\sigma_{\Omega,1})$ is $(0.19,0.033)\Gyr\inv$ in
both cases. Based on this information, a model with $\sigv =
0.365\kms$ and $\mu_\Omega = 6$ provides a good fit to the simulation
data. The distribution of frequency offsets in the two-dimensional
space perpendicular to the principal eigenvector of $V_\veco$ is
modeled as a zero-mean Gaussian with variance matrix given by the
projection of $V_\veco$ onto this space.

Because we expect the typical $\Delta\veco$ to scale with the velocity
dispersion of the progenitor, I conjecture that a constant
$\mu_\Omega$ can be used for modeling cold tidal streams of any
(small-ish) velocity dispersion. Further modeling of simulated
streams, however, is necessary to check this and to find the optimizal
value of $\mu_\Omega$. Fixing $\mu_\Omega$ leaves a single free
parameter \sigv\ for modeling the frequency offset distribution of a
tidal stream. In modeling to observed data, this parameter needs to be
fit and we expect it to be proportional to the velocity dispersion of
the progenitor, although this proportionality needs to be checked more
carefully (but see Figure 3 in \citealt{Sanders13a}, which shows that
the size of the frequency distribution scales as mass$^{1/3}$, or
approximately as the velocity dispersion). Similarly, the initial
angle distribution, modeled here as an isotropic Gaussian, has a
charactaristic spread $\sigma_{\theta}$ which I set here to $0.003$
($=\sigv/[122\kms]$), and we likewise expect this spread to scale with
the velocity dispersion of the progenitor.

The fiducial model for a tidal stream in frequency--angle coordinates
proposed here is therefore described by essentially two free
parameters in addition to the progenitor's phase-space position: the
progenitor's velocity dispersion \sigv\ and the disruption time
$t_d$. As shown below, these two parameters are the most important in
determining the track of the stream. More sophisticated models can
leave $\mu_\Omega$ and $\sigma_{\theta}$ as free parameters as
well. In what follows, I will fix these to the values given in the
previous paragraph. It must be stressed, however, that all of these
parameters here have been fixed ``by-eye'', and better fits might be
possible using more quantitative fit procedures.

@@

First, we compute the distribution of $\Delta \opar$, the frequency
offset between stream members and the progenitor given the angle
offset $\Delta \apar$ as follows (considering $\Delta \opar$ to be
positive)
\begin{equation}\label{eq:poparapargeneral}
\begin{split}
  p(\Delta \opar|\Delta \apar) & = \int \dd \ts\,p(\Delta \opar,\ts|\Delta \apar)\, \\
  & \propto \int \dd \ts \,p(\Delta \apar | \Delta \opar,\ts)\,p(\Delta \opar,\ts)\,,\\
  & \approx \int \dd \ts \,\delta(\Delta \apar - \Delta \opar\,\ts)\,p(\Delta \opar,\ts)\,,\\
  & = \frac{1}{\Delta \opar}\,p(\Delta \opar,\frac{\Delta \apar}{\Delta \opar})\,,
\end{split}
\end{equation}
where in the penultimate step I have approximated the initial angle
distribution as a delta function (as the initial angle offset is small
with respect to the final offset). This is a fully general
expression. For the fiducial stream model, this can be simplified to
\begin{equation}\label{eq:poparapar}
\begin{split}
  p(\Delta \opar|\Delta \apar) \propto &\ \frac{1}{\Delta \opar}\,p(\Delta \opar)\,,\qquad 0 < \frac{\Delta \apar}{\Delta \opar} < t_d\,.\\
  & 0\,,\qquad \mathrm{otherwise}\,.
\end{split}
\end{equation}
\Eqnname~(\ref{eq:poparapar}) shows why I chose to model $p(\Delta
\opar)$ as a Gaussian multiplied with $|\Delta \opar|$. $p(\Delta
\opar|\Delta \apar)$ is then a Gaussian with mean $\Delta \Omega^m$
and variance $\sigma^2_{\Omega,1}$, for $\Delta \opar > \Delta \apar /
t_d$, and zero otherwise. The mean and variance of such a Gaussian is
straightforward to calculate in terms of error functions. The top left
panel of \figurename~\ref{fig:gd1_apar} shows the distribution of
$|\Delta \opar|$ versus $|\Delta \apar|$ (absolute value in order to
show the trailing and leading stream together) for the simulated
stream, as well as the mean and dispersion calculated from the
fiducial model. The dashed line shows $\Delta \apar = \Delta \Omega^m
\,t_d$, approximately the angle where the finite disruption time
starts to influence the mean $\Delta \opar$ of the stream. The finite
disruption time, $\approx 4.5\Gyr$ in this case, of a tidal stream
means that for stream members to have reached very large angle
differences with respect to the progenitor, they must have been
removed at large frequency differences. Around $\Delta \Omega^m
\,t_d$, average stream stars, \ie, those removed with the average
frequency offset, do not have a large enough offset to reach large
angle offsets. Therefore, even though the average frequency offset
does not change over time (see \figurename~\ref{fig:gd1_times}), the
average frequency offset, and therefore the average orbit, does change
with distance from the progenitor. It is in this sense that streams do
not follow a single orbit. This is discussed further in
\sectionname~\ref{sec:discussorbit}.

@@

We can calculate $p(\ts | \Delta \apar)$ as follows
\begin{equation}\label{eq:pdtapar}
\begin{split}
  p(\ts | \Delta \apar) & = \int \dd \Delta \opar\,p(\ts,\Delta \opar|\Delta \apar)\,,\\
  & = \int \dd \Delta \opar\,p(\ts|\Delta \opar,\Delta \apar)\,p(\Delta \opar|\Delta \apar)\,,\\
  & = p\left(\Delta \opar = \frac{\Delta \apar}{\ts}\right)\,\left|\frac{\Delta \apar}{(\ts)^2}\right|\,,
\end{split}
\end{equation}
where in the last step I have again approximated the initial
distribution of parallel angle offsets as a delta function. The first
factor in this equation is given by the expression given in
\equationname~(\ref{eq:poparapargeneral}) (in general) or in
\equationname~(\ref{eq:poparapar}) for the fiducial model. The lower
right panel of \figurename~\ref{fig:gd1_apar} shows the distribution
of stripping times for members of the simulated stream as well as the
mean and dispersion calculated from
\equationname~(\ref{eq:pdtapar}). Similar to the distribution of
$\Delta \opar$ as a function of $\Delta \apar$, the distribution of
$\ts$ has a kink at the angle offset that can only be reached by stars
that have to have been stripped at large frequency offsets (that is,
larger than average).

We can now also calculate $p(\Delta \aperp|\Delta \apar)$:
\begin{equation}
  p(\Delta \aperpii|\Delta \apar) = \int \dd \ts \,\dd \Delta
  \operpii \,p(\Delta \aperpii|\Delta \operpii,\ts)\,p(\Delta \operpii|\ts,\Delta \apar)\,p(\ts|\Delta \apar)\,,
\end{equation}
where $ii=2,3$ for the first and second perpendicular direction. For
the fiducial model, this simplifies to
\begin{equation}\label{eq:paperpapar}
  p(\Delta \aperpii|\Delta \apar) = \int \dd \ts \,\mathcal{N}\left(\Delta \aperpii|0,\sigma_{\theta}^2+(\ts\,\sigma_{\Omega,ii})^2\right)\,p(\ts|\Delta \apar)\,,
\end{equation}
where $\sigma_{\Omega,ii}$ is the model frequency dispersion in the
perpendicular direction (corresponding to the middle and smallest
eigenvalue of $V_\veco$). I have \emph{not} approximated the initial
angle-offset distribution as a delta function here, because its width
is a significant fraction of that of the the final
perpendicular-angle-offset distribution. In the fiducial model, this
distribution is the only one of those considered in this section that
requires an explicit integration. However, we can approximate the
distribution by using the mean $\ts(\Delta \apar)$ rather than
integrating over $\ts$. The lower left panel of
\figurename~\ref{fig:gd1_apar} shows the distribution of perpendicular
angle offsets and the grayscale bands show the dispersion calculated
using \equationname~(\ref{eq:paperpapar}); the white dashed line shows
the approximate estimate obtained using the mean stripping time rather
than integrating over it. This simple estimate agrees well with the
exact calculation.
